{ Das CrypTool-Skript: \\ Kryptographie, Mathematik und mehr} \\ \vspace{30pt} {\large Hintergrundmaterial und Zusatzinformationen\\%*[-8pt] zum freien E-Learning-Programm CrypTool\\ (9. Auflage -- zu Version 1.4.20)} \vspace{60pt}

### 2009-07-20 script/de/chapters/paper_and_pencil.tex
CT script, paper and pencil: made some general corrections; improved the code of the all Sage samples (de+en) in this chapter (more educational)

% $Id$
% ............................................................................
%             V E R S C H L U E S S E L U N G S V E R F A H R E N
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% "`xx"'
\newpage
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% corrected by C. Esslinger, June 2005 
%------------------------------------------------------------------------------
\hypertarget{Kapitel_PaperandPencil}{}

\chapter{Papier- und Bleistift-Verschlüsselungsverfahren}
\label{Kapitel_PaperandPencil}
(Christine Stötzel, April 2004; Updates B.+C. Esslinger, Juni 2005; Updates B. Esslinger, Juli 2009)
\index{Papier- und Bleistiftverfahren}

@@

\item {\bf Gartenzaun}\footnote{In CrypTool
   kann man dieses Verfahren über den Menüeintrag {\bf Ver-/Entschlüsseln
   \textbackslash{} Symmetrisch (klassisch) \textbackslash{} Permutation} 
   abbilden: für einen Gartenzaun mit 2 Zeilen gibt man als Schlüssel
   "`B,A"' ein und lässt ansonsten die Standardeinstellungen (nur 1
   Permutation, in der man zeilenweise ein- und spaltenweise ausliest). 
   Mit dem Schlüssel "`A,B"' würde man man das Zick-zack-Muster unten so 
   beginnen, dass der 1. Buchstabe in der ersten statt in der 2. Zeile steht.}
   \cite{Singh2001}\index{Gartenzaun-Verschlüsselung}:
                        % Kein Blank VOR "\index", sonst ist der
                        % Doppelpunkt nicht direkt nach dem Fussnotenindex.
   Die Buchstaben des Klartextes werden abwechselnd in zwei (oder mehr) 
   Zeilen geschrieben, so dass ein Zickzack-Muster entsteht. 
   Dann werden die Zeichen zeilenweise nacheinander gelesen.\\
   Dieses Verfahren ist eher eine Kinderverschlüsselung.

@@

\item {\bf Skytale von Sparta}\footnote{%
   Das Ergebnis dieses Verschlüsselungsverfahrens entspricht dem einer 
   einfachen Spaltentransposition.  
   In CrypTool kann man dieses Verfahren über den Menüeintrag {\bf 
   Ver-/Entschlüsseln \textbackslash{} Symmetrisch (klassisch) 
   \textbackslash{} Permutation} abbilden: Für die Skytale braucht man in
   der Dialogbox nur die erste Permutation. Darin gibt man bei z.B. 4 
   Kanten als Schlüssel  "`1,2,3,4"'  ein. Dies wäre so, als würde man
   den Text in 4-er Blöcken waagrecht in eine Tabelle schreiben und 
   senkrecht auslesen. 
   Weil der Schlüssel aufsteigend geordnet ist, bezeichnet man die Skytale
   auch als identische Permutation. Weil das Schreiben und Auslesen nur 
   einmal durchgeführt wird, als einfache (und nicht als doppelte)
   Permutation.}
   \cite{Singh2001}\index{Skytale}%
   : 
   Dieses Verfahren wurde wahrscheinlich das erste Mal um 600 v.Chr.
   benutzt und es wurde von dem griechischen Schriftsteller und Philosophen
   Plutarch (50-120 v.Chr.) zuerst beschrieben.\\
   Um einen Holzstab wird ein Streifen Papier o.ä. gewickelt. Dann wird
   darauf zeilenweise der Klartext geschrieben. 
   Wickelt man den Streifen ab, hält man den Geheimtext in den Händen.

@@

@@

Im Folgenden werden einige der klassischen Verfahren mit dem Computer-Algebra-System Sage implementiert. Der Code ist lauffähig und getestet mit Sage Version 4.1.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-de}
\caption{Struktur und Namenskonventionen in den Sage Programmbeispielen} 
\label{XXX}
\end{center}
\end{figure}

\subsection{Transposition ciphers using Sage}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains our plaintext and ciphertext
spaces. Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by one character, with wrap
around. That is the encryption process. The decryption process is
shifting to the left by $14 - 1 = 13$ characters in order to recover
the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher of block length 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # our encryption/decryption key
sage: key = [2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1]
sage:
sage: # encrypt a plaintext
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
BCDEFGHIJKLMNA
sage:
sage: # decrypt ciphertext
sage: keyInv = [14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}

The encryption key can also be such that it shifts each letter of the
plaintext to the left by two characters, with wrap around. Then we
need to shift each letter of the ciphertext to the left by $14 - 2 =
12$ characters in order to obtain the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # encrypt by shifting to the left by 2 letters
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}

As a more complicated example, we can use a permutation in the
encryption and decryption processes. If the block length of our
transposition cipher is $n$, then the encryption/decryption
key must be a permutation on $n$ symbols. The following
example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. So each call to
\verb!random_key()! produces a different key for encryption, which
affects the ciphertext. Note that your results may be different from
the following example.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage:
sage: # encryption
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: P = T.encoding("a b c d e f g h i j k l m n")
sage: C = T.enciphering(key, P); C
BGIEJNAMCLDHKF
sage:
sage: # decryption
sage: T.deciphering(key, C)
ABCDEFGHIJKLMN
sage: T.deciphering(key, T.enciphering(key, P))
ABCDEFGHIJKLMN
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection{Substitution ciphers}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\verb!SubstitutionCryptosystem!. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results. First, we perform the encryption process with a
random key.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage:
sage: # substitution cipher with random key
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
CMPGOWFIJRESXNYBKUDALQHTVZ
sage:
sage: # encryption
sage: msg = A.encoding("Substitute this with something else better."); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: S.enciphering(key, msg)
DLMDAJALAOAIJDHJAIDYXOAIJNFOSDOMOAAOU
\end{Verbatim}

>>>>

% $Id$
% ............................................................................
%             V E R S C H L U E S S E L U N G S V E R F A H R E N
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% "`xx"'
\newpage
\hypertarget{Kapitel_PaperandPencil}{}

\chapter{Papier- und Bleistift-Verschlüsselungsverfahren}
\label{Kapitel_PaperandPencil}
(Christine Stötzel, April 2004; Updates B.+C. Esslinger, Juni 2005; Updates Minh Van Nguyen und B. Esslinger, Juli 2009)
\index{Papier- und Bleistiftverfahren}

@@

\item {\bf Gartenzaun}\footnote{%
   Dieses Verfahren kann direkt in CrypTool\index{CrypTool} mit dem Menüpunkt 
   {\bf Ver-/Entschlüsseln \textbackslash{} Symmetrisch (klassisch) \textbackslash{}
   Skytale / Gartenzaun} abgebildet werden.
   Man kann diese Matheode auch simulieren über den Menüeintrag {\bf Ver-/Entschlüsseln
   \textbackslash{} Symmetrisch (klassisch) \textbackslash{} Permutation}:
   für einen Gartenzaun mit 2 Zeilen gibt man als Schlüssel
   "`B,A"' ein und lässt ansonsten die Standardeinstellungen (nur 1
   Permutation, in der man zeilenweise ein- und spaltenweise ausliest). 
   Mit dem Schlüssel "`A,B"' würde man man das Zick-zack-Muster unten so 
   beginnen, dass der 1. Buchstabe in der ersten statt in der 2. Zeile steht.}
   \cite{Singh2001}\index{Gartenzaun-Verschlüsselung}:
                        % Kein Blank VOR "\index", sonst ist der
                        % Doppelpunkt nicht direkt nach dem Fussnotenindex.
   Die Buchstaben des Klartextes werden abwechselnd in zwei (oder mehr) 
   Zeilen geschrieben, so dass ein Zickzack-Muster entsteht. 
   Dann werden die Zeichen zeilenweise nacheinander gelesen.\\
   Dieses Verfahren ist eher eine Kinderverschlüsselung.

@@

\item {\bf Skytale von Sparta}\footnote{%
   Dieses Verfahren kann direkt in CrypTool\index{CrypTool} mit dem Menüpunkt 
   {\bf Ver-/Entschlüsseln \textbackslash{} Symmetrisch (klassisch) \textbackslash{}
   Skytale / Gartenzaun} abgebildet werden.
   Da dieses Verschlüsselungsverfahrens ein Spezialfall der einfachen
   Spaltentransposition ist, kann man es in CrypTool\index{CrypTool} simulieren
   über den Menüeintrag {\bf Ver-/Entschlüsseln \textbackslash{} Symmetrisch (klassisch) 
   \textbackslash{} Permutation}: Für die Skytale braucht man in
   der Dialogbox nur die erste Permutation. Darin gibt man bei z.B. 4 
   Kanten als Schlüssel  "`1,2,3,4"'  ein. Dies wäre so, als würde man
   den Text in 4-er Blöcken waagrecht in eine Tabelle schreiben und 
   senkrecht auslesen. 
   Weil der Schlüssel aufsteigend geordnet ist, bezeichnet man die Skytale
   auch als identische Permutation. Weil das Schreiben und Auslesen nur 
   einmal durchgeführt wird, als einfache (und nicht als doppelte)
   Permutation.}
   \cite{Singh2001}\index{Skytale}%
   : 
   Dieses Verfahren wurde wahrscheinlich das erste Mal um 600 v.Chr.
   benutzt und es wurde von dem griechischen Schriftsteller und Philosophen
   Plutarch (50-120 v.Chr.) zuerst beschrieben.\\
   Um einen Holzstab wird ein Streifen Papier o.ä. gewickelt. Dann wird
   darauf zeilenweise der Klartext geschrieben. 
   Wickelt man den Streifen ab, hält man den Geheimtext in den Händen.

@@

Im Folgenden werden einige der klassischen Verfahren mit dem Open-Source Computer-Algebra-System
Sage implementiert. Der Code ist lauffähig und getestet mit Sage Version 4.1.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-de}
\caption{Namenskonventionen in den Sage-Programmbeispielen zu Verschlüsselungsverfahren} 
\label{XXX}
\end{center}
\end{figure}

\newpage
\subsection{Transposition ciphers}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the spave of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by two characters, with wrap
around. That is the encryption process. The decryption process is
shifting each letter of the ciphertext to the left by $14 - 2 = 12$
characters in order to recover the plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case to upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key explicitely given)}
\end{sagecode}

\newpage
The next sample is also a simple column transposition. But now the code is a little bit more automated in generating the keys from the shift parameter.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key constructed with "range")}
\end{sagecode}

\newpage
As a more complicated example, we can use a permutation in the
encryption and decryption processes. If the block length of our
transposition cipher is $n$, then the encryption/decryption
key must be a permutation on $n$ symbols. The following
example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. So each call to
\verb!random_key()! produces a different key for encryption, which
affects the ciphertext. Note that your results may be different from
the following example.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
ssage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Simple Column Transposition with randomly generated (permutation) key}
\end{sagecode}

\newpage
The following example shows the key space of a transpositon cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage: T.key_space()
Symmetric group of order 14! as a permutation group
sage: # Remark: The key space is not quite correct as also permutations shorter than keylen are counted.
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # enciphering in one and in two steps
sage: C   = T.enciphering(key, msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: enc = T(key); enc.key()
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: C = enc(msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: # deciphering
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
\end{Verbatim}
\caption{Simple Column Transposition (showing the key\_space)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution ciphers}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCrypto\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results. First, we perform the encryption process with a
random key.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #.# We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Caesar cipher}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitely given, step-by-step approach)}
\end{sagecode}

\newpage
The next sample does the same, but the code is more sophisticated/automated/variable.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; keys generated)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution with symbols}

Here is a Sage example where the symbols are from the binary number
system. A monoalphabetic substitution cipher with a binary alphabet has very little
security: Because the plaintext/ciphertext alphabet has only the two elements
0 and 1, there are only two keys possible: (0 1) and (1 0).

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = BinaryStrings()
sage: # substitution cipher over the alphabet A; no keylen argument possible
sage: S = SubstitutionCryptosystem(A); S
Substitution cryptosystem on Free binary string monoid
sage: # to have a substitute for each symbol, key has always the length of the alphabet
sage: key = S.random_key(); key
10
sage: len(key)
2
sage: P = "Working with binary numbers."
sage: # encryption
sage: msg = A.encoding(P); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: C   = S.enciphering(key, msg); C
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110011101100101101001000110011110100011011000011011\
0111111001000110001010100100101001110110011010100011011000110011010001
sage: # decryption
sage: DC  = S.deciphering(key, C); DC
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001100010011010010110111001100001011100100111100100\
1000000110111001110101011011010110001001100101011100100111001100101110
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a binary alphabet}
\end{sagecode}

\newpage
However, our plaintext/ciphertext space  can be over a larger
alphabet, say the hexadecimal number system. Here is an example
using the hexadecimal number system as our substitution alphabet.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
2b56a4e701c98df3
sage: len(key)
16
sage: # Number of possible keys
sage: factorial(len(key))
20922789888000
sage: P   = "Working with a larger alphabet."
sage:
sage: msg = A.encoding(P); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: C   = S.enciphering(key, msg); C
47e375e9e1efe75277e17ae052eb52e8eb75e7e47552ebe872e0ebe5e47a5f
sage: DC  = S.deciphering(key, C); DC
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: msg == DC
True
sage:
sage: #.# Geht das auch mit den Methoden des Alphabets oder der Chiffren ?????
sage: # Conversion hex back to ASCII:
sage: # - AlphabeticStrings() and HexadecimalStrings() don't have according methods.
sage: # - So we used Python:
sage: #   - repr(DC) converts the StringMonoidElement DC to a string
sage: #     still containing hex symbols;
sage: #   - binascii.a2b_hex converts this hex string to an ascii string.
sage: import binascii
sage: DDC = binascii.a2b_hex(repr(DC)); DDC
'Working with a larger alphabet.'
sage:
sage: P == DDC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with a hexadecimal alphabet (and decoding in Python)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Vigen{\`e}re cipher}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: keylen = 14
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, keylen); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of period 14
sage: #.# V.key_space() lieferte keinen Fehler, aber auch kein Ergebnis.
sage:
sage: # alternative: given key: key = A('ABCDEFGHIJKLMN'); key
sage: key = V.random_key(); key
WSSSEEGVVAARUD
sage: len(key)
14
sage:
sage: # encoding
sage: P = "The Vigenere cipher is polyalphabetic."
sage: len(P)
38
sage: msg = V.encoding(P); msg     # alternative: msg = A.encoding(P); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage:
sage: # encryption [2 alternative ways (in two steps or in one): both work]
sage: # encrypt = V(key); encrypt
sage: # C = encrypt(msg); C
sage: C   = V.enciphering(key, msg); C
PZWNMKKIZRETCSDWJAWTUGTALGBDXWLAG
sage:
sage: # decryption
sage: DC  = V.deciphering(key, C); DC
THEVIGENERECIPHERISPOLYALPHABETIC
sage: msg == DC
True
\end{Verbatim}
\caption{Vigen{\`e}re cipher}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Hill cipher}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct a Hill cipher
sage: keylen = 19    # keylen = 3  # Alternative key length with non-random small key
sage: A = AlphabeticStrings()
sage: H = HillCryptosystem(A, keylen); H
Hill cryptosystem on Free alphabetic string monoid on A-Z of block length 19
sage:
sage: # To create key non-randomly, HKS is necessary [even H.key_space() is not enough].
sage: # HKS = H.key_space()
sage: # key = HKS([[1,0,1],[0,1,1],[2,2,3]]); key
sage:
sage: # Random key creation
sage: key = H.random_key(); key
[10  7  5  2  0  6 10 23 15  7 17 19 18  2  9 12  0 10 11]
[23  1  1 10  4  9 21  1 25 22 19  8 17 22 15  8 12 25 22]
[ 4 12 16 15  1 12 24  5  9 13  5 15  8 21 23 24 22 20  6]
[ 5 11  6  7  3 12  8  9 21 20  9  4 16 18 10  3  2 23 18]
[ 8 22 14 14 20 13 21 19  3 13  2 11 13 23  9 25 25  6  8]
[24 25  8 24  7 18  3 20  6 11 25  5  6 19  7 24  2  4 10]
[15 25 11  1  4  7 11 24 20  2 18  4  9  8 12 19 24  0 12]
[14  6  2  9 11 20 13  4 10 11  4 23 14 22 14 16  9 12 18]
[12 10 21  5 21 15 16 17 19 20  1  1 15  5  0  2 23  4 14]
[21 15 15 16 15 20  4 10 25  7 15  4  7 12 24  9 19 10  6]
[25 15  2  3 17 23 21 16  8 18 23  4 22 11 15 19  6  0 15]
[14 23  9  3 18 15 10 18  7  5 12 23 11  9 22 21 20  4 14]
[ 3  6  8 13 20 16 11  1 13 10  4 21 25 15 12  3  0 11 18]
[21 25 14  6 11  3 21  0 19 17  5  8  5  4  9  2 23 19 15]
[ 8 11  9 11 20 15  6  1  3 18 18 22 16 17  6  3 15 11  2]
[21 15  5 22  2  9  0  4 22 10  2 10 19 19 17 19  1 21  4]
[ 7 17  9  2 15  5 14  3  6  9 12 12 22 15  8  4 21 14 19]
[19 14 24 19  7  5 22 22 13 14  7 18 17 19 25  2  1 23  6]
[ 2  6 14 22 17  7 23  6 22  7 13 20  0 14 23 17  6  1 12]
sage:
sage: # encoding and encryption
sage: P = "Hill or matrix cipher uses matrix operations."
sage: len(P)
45
sage: # implementation requires: Length of msg is a multiple of matrix dimension (block_length)
sage: msg = H.encoding(P); msg
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
sage: len(msg)
38
sage:
sage: # encryption
sage: C  = H.enciphering(key, msg); C
CRWCKPRVYXNBRZTNZCTQWFWSDWBCHABGMNEHVP
sage:
sage: # decryption
sage: DC  = H.deciphering(key, C); DC
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
sage: msg == DC
True
sage:
sage: # alternative decryption using inverse matrix
sage: keyInv = H.inverse_key(key); keyInv
[ 6 23  1 23  3 12 17 22  6 16 22 14 18  3  1 10 21 16 20]
[18 23 15 25 24 23  7  4 10  7 21  7  9  0 13 22  5  5 23]
...
[10 11 12  6 11 17 13  9 19 16 14 24  4  8  5 16 18 20  1]
[19 16 16 21  1 19  7 12  3 18  1 17  7 10 24 21  7 16 11]
sage: DC     = H.enciphering(keyInv, C); DC
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
\end{Verbatim}
\caption{Hill cipher}
\end{sagecode}

@@
